---
title: "lab 16"
author: "Sophia Halling"
date: "2024-05-01"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Examples

### Libraries

```{r}
library(tidyverse)
library(knitr)
library(ggtree)
library(TDbook) #A Companion Package for the Book "Data Integration, Manipulation and Visualization of Phylogenetic Trees" by Guangchuang Yu (2022, ISBN:9781032233574).
library(ggimage)
library(rphylopic)
library(treeio)
library(tidytree)
library(ape)
library(TreeTools)
library(phytools)
library(ggnewscale)
library(ggtreeExtra)
library(ggstar)
```

### Read in files

#### Mag table

```{r}
NEON_MAGs <- read_csv("data/NEON/GOLD_Study_ID_Gs0161344_NEON_2024_4_21.csv") %>% 
  # remove columns that are not needed for data analysis
  select(-c(`GOLD Study ID`, `Bin Methods`, `Created By`, `Date Added`, `Bin Lineage`)) %>% 
  # create a new column with the Assembly Type
  mutate("Assembly Type" = case_when(`Genome Name` == "NEON combined assembly" ~ `Genome Name`,
                            TRUE ~ "Individual")) %>% 
  mutate_at("Assembly Type", str_replace, "NEON combined assembly", "Combined") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "d__", "") %>%  
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "p__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "c__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "o__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "f__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "g__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "s__", "") %>%
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";", remove = FALSE) %>% 
  mutate_at("Domain", na_if,"") %>% 
  mutate_at("Phylum", na_if,"") %>% 
  mutate_at("Class", na_if,"") %>% 
  mutate_at("Order", na_if,"") %>% 
  mutate_at("Family", na_if,"") %>% 
  mutate_at("Genus", na_if,"") %>% 
  mutate_at("Species", na_if,"") %>% 
  
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  # Get rid of the the common string "S-comp-1"
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

#### Metagenome table

```{r}
NEON_metagenomes <- read_tsv("data/NEON/exported_img_data_Gs0161344_NEON.tsv") %>% 
  select(-c(`Domain`, `Sequencing Status`, `Sequencing Center`)) %>% 
  rename(`Genome Name` = `Genome Name / Sample Name`) %>% 
  filter(str_detect(`Genome Name`, 're-annotation', negate = T)) %>% 
  filter(str_detect(`Genome Name`, 'WREF plot', negate = T)) 
```

```{r}
NEON_metagenomes <- NEON_metagenomes %>% 
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  # Get rid of the the common string "-comp-1"
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

#### NEON chemistry data

```{r}
NEON_chemistry <- read_tsv("data/NEON/neon_plot_soilChem1_metadata.tsv") %>% 
  # remove -COMP from genomicsSampleID
  mutate_at("genomicsSampleID", str_replace, "-COMP", "") 
```

#### Join data

```{r}
NEON_MAGs_metagenomes_chemistry <- NEON_MAGs %>% 
  left_join(NEON_metagenomes, by = "Sample Name") %>% 
  left_join(NEON_chemistry, by = c("Sample Name" = "genomicsSampleID")) %>% 
  rename("label" = "Bin ID")
```


### Trees

```{r}
tree_arc <- read.tree("data/NEON/gtdbtk.ar53.decorated.tree")
tree_bac <- read.tree("data/NEON/gtdbtk.bac120.decorated.tree")
```

#### Subtree

```{r}
# Make a vector with the internal node labels
node_vector_bac = c(tree_bac$tip.label,tree_bac$node.label)

# Search for your Phylum or Class to get the node
grep("Bacteroidota", node_vector_bac, value = TRUE)
```

```{r}
match(grep("Bacteroidota", node_vector_bac, value = TRUE), node_vector_bac)
```

```{r}
tree_bac_preorder <- Preorder(tree_bac)
tree_Bacteroidota <- Subtree(tree_bac_preorder, 2507)
```

### Filter NEON data

```{r}
NEON_MAGs_Bacteroidota <- NEON_MAGs_metagenomes_chemistry %>% 
  filter(Phylum == "Bacteroidota") 
```


### Tree graphs

#### Rectangular

```{r}
ggtree(tree_Bacteroidota) +
  geom_tiplab(size=3) +
  xlim(0,20)
```

#### Circular

```{r}
ggtree(tree_Bacteroidota, layout="circular") + 
  geom_tiplab(aes(angle=angle))+
    theme_tree() +
    xlim(0,20)
```
#### Circular with taxonomic highlight

```{r}
ggtree(tree_bac, layout="circular") +
    geom_hilight(node=2507, fill="steelblue", alpha=.6) 
```

#### circular with taxonomic label

```{r}
ggtree(tree_bac, layout="circular", branch.length="none") +
    geom_hilight(node=2507, fill="steelblue", alpha=.6) +
    geom_cladelab(node=2507, label="Bacteroidota", align=TRUE,  
                  offset = 0, textcolor='steelblue', barcolor='steelblue')
```

#### circular with taxonomy collapsed

```{r}
ggtree(tree_bac, layout="circular", branch.length="none") %>% 
  collapse(node=2507) + 
  geom_point2(aes(subset=(node==55)), shape=23, size=5, fill='steelblue') +
  geom_cladelab(node=55, label="Bacteroidota", align=TRUE,  
                  offset = 2, textcolor='steelblue')
```

#### circular with taxonomy collapsed into a triangle

```{r}
p <- ggtree(tree_bac, layout="circular", branch.length="none")
scaleClade(p, 2507, .2) %>% collapse(2507, 'min', fill="steelblue")  +
  geom_cladelab(node=2507, label="Bacteroidota", align=TRUE,  
                  offset = 0.2, textcolor='steelblue')
```

#### adding data to the tree

```{r}
ggtree(tree_Bacteroidota, layout="circular")  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tiplab(size=2, hjust=-.1) +
  xlim(0,20) +
  geom_point(mapping=aes(color=Class)) 
```

```{r}
ggtree(tree_Bacteroidota, layout="circular")  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tiplab(size=2, hjust=-.1) +
  xlim(0,20) +
  geom_point(mapping=aes(color=Class, shape = `Assembly Type`))
```

```{r}
ggtree(tree_Bacteroidota)  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tiplab(size=2, hjust=-.1) +
  xlim(0,20) +
  geom_point(mapping=aes(color=`Ecosystem Subtype`)) 
```

#### adding a gradient

```{r}
ggtree(tree_Bacteroidota)  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_tiplab(size=2, hjust=-.1) +
  xlim(0,20) +
  geom_point(mapping=aes(color=`Ecosystem Subtype`)) 
```

#### circular tree with data

```{r}
ggtree(tree_Bacteroidota, layout="circular")  %<+%
  NEON_MAGs_metagenomes_chemistry + 
  geom_point2(mapping=aes(color=`Ecosystem Subtype`, size=`Total Number of Bases`))
```

